---
title: "ANOVA_in_RSTUDIO"
author: "Trevor Harrington"
format: html
editor: visual
keep-md: TRUE
prefer-html: TRUE
code-fold: TRUE
---

```{r echo= FALSE}
library(readr)
library(dplyr)
library(ggplot2)
```

```{r}
datum <- read_csv("C:/School/23SPDAY/FieldMethodsandTech/Practice Data/linear regression/anova/Class Activity 9.csv",
                             show_col_types = FALSE)
names(datum)[names(datum) == "Land use"] <- "land_use"
list(datum)
```

## ANOVA analysis in R Studio

-   ANOVA can be used to **compare the means between three or more independent groups**

## ANOVA Testing: When and Why to Use ANOVA

### What is ANOVA

-   ANOVA (Analysis of Variance) is a statistical technique used to **compare means between three or more groups**.

-   ANOVA tests whether the differences between group means are statistically significant or simply due to chance.

-   The basic idea behind ANOVA is to partition the total variability in a dataset into two parts: variability between groups and variability within groups.

    -   If the variability between groups is much greater than the variability within groups, then it is likely that the group means are different and that the observed differences are not due to chance.

-   **F- Values** describe the **ratio of the variance between groups to the variance within groups**. It indicates the degree of variation in the dependent variable (in this case, the amount of nitrogen generated) that is explained by the independent variable (in this case, the land use types).

### Short Summary of the Math

------------------------------------------------------------------------

## Viewing the data

-   always start with viewing a table what the data contains and generating an exploratory graph to show the correlation you hope to predict

### Generating a Data Summary

```{r}
summary(datum)
```

### Producing a plot

```{r}
datum %>% 
  ggplot(aes(x = land_use, y = Nitrogen, fill = land_use)) +
  geom_boxplot() +
  labs(x = "Land Use", y = "Nitrogen", title = "Nitrogen by Land Use") +
  scale_fill_manual(values = c("Agriculture" = "lightblue", "Forest" = "darkgreen", "Developed" = "gray")) +
  theme_classic()

```

::: {.callout-note .div style="border: 1px solid #ccc;   background-color: #f2f2f2;   color: #333;   padding: 10px;   margin: 10px 0;"}
## Observations

This graph suggests that **land use for developement** as a source of nitrogen seems to be significantly greater then both agriculture and forest. It also seems that the relationship between forest and agriculture is also significant.
:::

### Running an ANOVA to test the significant of this comparison

```{r}
results = aov(Nitrogen~as.factor(land_use), data= datum)

summary(results)

```

```{r}
TukeyHSD(results)
```

::: {.callout-note .div style="border: 1px solid #ccc;   background-color: #f2f2f2;   color: #333;   padding: 10px;   margin: 10px 0;"}
## Conclusion

The ANOVA test showed a significant difference in the mean amount of nitrogen generated between land uses (developed, agriculture, and forest) with a large F-value (153.6) suggesting a much greater variation between groups, then within them. and a p-value \< 0.05. The Tukey post-hoc analysis revealed that the mean amount of nitrogen generated in developed land use (mean = 2.74) was significantly higher than that of agriculture (mean = 2.26) and forest (mean = 2.59) land use with p-values \< 0.001 and \< 0.0001, respectively. On the other hand, the mean amount of nitrogen generated in forest land use was significantly lower than that of developed and agriculture land use with p-values \< 0.0001 and 0.0007, respectively. However, there was no significant difference in the mean amount of nitrogen generated between agriculture and forest land use (p = 0.70).
:::
