---
title: "Assignment 5"
author: "Trevor Harrington"
format: html
editor: visual
execute: 
  keep-md: TRUE
  code-fold: TRUE
---

# Assignment 5: Applying t-test and ANOVA

```{r}
library(tidyr)
library(ggplot2)
library(kableExtra)
library(readr)
```

## 5-1: Comparing the soil carbon concentration in traditional vs. regenerative agriculture - [t-test]{.underline}

```{r}
datum <- read_csv("~/R/Field_methods/Week 9 Assignments/HW 5-dataset1.csv",
                  show_col_types = FALSE)
datum %>%
kable(digits = 4) %>%
  kable_styling(bootstrap_options = "striped", "hover")
```

```{r}
datum %>%
  summary()
```

```{r}
library(ggplot2)

ggplot(data = datum, aes(x = as.factor(Agriculture.Practice), y = Carbon, fill = as.factor(Agriculture.Practice))) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "black") +
  labs(x = "Agriculture Practice", y = "Carbon", fill = "Agriculture Practice", title = "Carbon by Agriculture Practice") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))

```

```{r}
results=t.test(Carbon~Agriculture.Practice,data=datum,var.equal=FALSE)

results 
```

-   This model does not find a significant difference between the soil concentration in regenerative vs traditional agricultural practices. 95% confidence interval includes 0 indicating that there is not enough evidence to reject the null hypothesis of no difference in means between the two groups at a significance level of 0.05. ( t=0.9454, p=0.3621)

## 5-1: Comparing Four Bobwhite Quail Habitat types and population density - ANOVA

```{r}
datum <- read_csv("~/R/Field_methods/Week 9 Assignments/HW 5-dataset2.csv",
                  show_col_types = FALSE)
datum %>%
  kable() %>%
  kable_styling(bootstrap_options = "striped", "hover")
```

```{r}
datum <- read_csv("~/R/Field_methods/Week 9 Assignments/HW 5-dataset2.csv",
                  show_col_types = FALSE)
summary(datum) %>%
  kable() %>%
  kable_styling(bootstrap_options = "striped", "hover")
```

```{r}
datum %>% 
  ggplot(aes(x = Habitat, y = QuailDens, fill = Habitat)) +
  geom_boxplot() +
  labs(x = "Habitat", y = "Bobwhite Quail Dens", title = "distribution of Bobwhite Quail Dens by habitat") +
  scale_fill_manual(values = c("OpenPine" = "lightblue", "ClosedPine" = "maroon", "Prarie" = "darkgreen", "Ag" = "orange")) +
  theme_classic()
```

hypothesis: From this graph, it looks like there will be no significant statistical difference between Agriculture and Closed pine habitats, or between Open Pine and Prairie habitats . However, between these two groups it seems likely that there will be a statistically significant difference in mean population density.

```{r}
results = aov(QuailDens~as.factor(Habitat), data= datum)

summary(results)

```

```{r}
TukeyHSD(results)
```

-   **OpenPine-Ag: The estimated difference in population density between the OpenPine and Agriculture habitats is -0.41 individuals per hectare, with a 95% confidence interval ranging from -0.77 to -0.04 individuals per hectare (p = 0.02), indicating a statistically significant difference.**

-   Prairie-Ag: The estimated difference in population density between the Prairie and Agriculture habitats is -0.38 individuals per hectare, with a 95% confidence interval ranging from -0.74 to -0.01 individuals per hectare (p = 0.04), indicating a statistically significant difference.

-   **ClosedPine-Ag: The estimated difference in population density between the ClosedPine and Agriculture habitats is -0.03 individuals per hectare, with a 95% confidence interval ranging from -0.39 to 0.34 individuals per hectare (p = 0.99), indicating that there is no statistically significant difference.**

-   OpenPine-ClosedPine: The estimated difference in population density between the OpenPine and ClosedPine habitats is -0.38 individuals per hectare, with a 95% confidence interval ranging from -0.75 to -0.02 individuals per hectare (p = 0.04), indicating a statistically significant difference.

-   Prairie-ClosedPine: The estimated difference in population density between the Prairie and ClosedPine habitats is -0.35 individuals per hectare, with a 95% confidence interval ranging from -0.72 to 0.01 individuals per hectare (p = 0.06), which is not statistically significant (i.e., p \> 0.05).

-   Prairie-OpenPine: The estimated difference in population density between the Prairie and OpenPine habitats is 0.03 individuals per hectare, with a 95% confidence interval ranging from -0.34 to 0.39 individuals per hectare (p = 0.99), indicating that there is no statistically significant difference.

the hypothesis is partially supported. The comparison between Agriculture and Closed Pine habitats did not show a significant difference in population density, which is consistent with the hypothesis. However, the comparison between Open Pine and Prairie habitats did not show a significant difference in population density, which is not consistent with the hypothesis. In addition, the comparison between Prairie and Closed Pine habitats had a p-value of 0.06, which is marginally above the threshold of 0.05 for statistical significance, suggesting a possible difference in population density that should be further investigated. Therefore, while the hypothesis correctly predicted one comparison, it was not entirely accurate for the other comparisons.
