---
title: "RegressioninRStudio"
author: "Trevor Harrington"
format: html
editor: visual
execute: 
  keep-md: TRUE
  code-fold: TRUE
---

```{r}
library(readr)
Dataset_1 <- read_csv("~/R/Field_methods/Dataset 1.csv")
View(Dataset_1)

Dataset_2 <- read_csv("~/R/Field_methods/Dataset 2.csv")
View(Dataset_2)

Dataset_3 <- read_csv("~/R/Field_methods/Dataset 2.csv")
View(Dataset_3)

```

## 1. Wolf Spiders vs. Prey Spiders

```{r}

plot(WolfSpiders~PreySpiders, data = Dataset_1)

title(main = "Wolf vs Prey Spiders")

results = lm(WolfSpiders~PreySpiders, data = Dataset_1)

abline(results)

```

```{r}
results = lm(WolfSpiders~PreySpiders, data = Dataset_1)

summary(results)
```

### Conclusion

**For each 1 spider/m2 increase in prey spiders, we observed a 2.05 spider/m2 increase in wolf spiders (p = 3.77x10\^-6; r\^2 = 0.419).**

## 2. Relationship Between Elevation and Size of Lotus Flowers

```{r}

plot(Size~Elevation, data = Dataset_2)
title(main = "Elevation vs. Lotus Flower Size")
results = lm(Size~Elevation, data = Dataset_2)
abline(results)
```

```{r}
results = lm(Size~Elevation, data = Dataset_2)
summary(results)
```

### Conclusion

**For each 1000 meter increase in elevation, we observed a 1.01cm decrease in lotus flower size (p = 2 x 10\^-16; r\^2 = 0.1459).**

## 3. Relationship between Foraging Time and Bird of Prey Presence in Colombian Ground Squirrels

```{r}

plot(Predators~Time, data = Dataset_3)
title(main = "Time Spent Foraging vs. Birds of Prey Sightings")
results = lm(Predators~Time, data = Dataset_3)
abline(results)
```

```{r}
results = lm(Predators~Time, data = Dataset_3)
summary(results)
```

### Conclusion

**For each 1 birds of prey increase measured the preceding day, we observed a 2.81 minute/day increase in time spent foraging by Colombian Squires (p = 0.106; r\^2 = 0.0117).**
